---
title: "Land Cover Change Detection for Floodplains"
output: bookdown::html_document2
vignette: >
  %\VignetteIndexEntry{Land Cover Change Detection for Floodplains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the drift pipeline using a small floodplain reach
on Neexdzii Kwa (Upper Bulkley River) in northern BC. We compare Esri IO LULC
land cover across 2017, 2020, and 2023 to track vegetation and land use change
in the riparian zone.

The AOI polygon was delineated using the
[flooded](https://github.com/NewGraphEnvironment/flooded) package, which
identifies floodplain extents from DEMs and stream networks.

The example data ships with the package — no STAC queries or database
connections needed.

## Load Data

```{r load-data}
library(drift)
library(terra)
library(sf)

# AOI polygon (floodplain delineated via flooded package)
aoi <- sf::st_read(
  system.file("extdata", "example_aoi.gpkg", package = "drift"),
  quiet = TRUE
)

# IO LULC rasters for 3 years
years <- c(2017, 2020, 2023)
rasters <- lapply(years, function(yr) {
  terra::rast(system.file("extdata", paste0("example_", yr, ".tif"),
                          package = "drift"))
})
names(rasters) <- years
```

## Classify

Apply IO LULC class names and colors from the shipped class table.

```{r classify}
classified <- dft_rast_classify(rasters, source = "io-lulc")

# Check factor levels
terra::levels(classified[["2020"]])[[1]]
```

## Classified Rasters

The figure below shows the three classified time steps side by side.

```{r plot-classified, fig.cap = "Classified land cover for the Neexdzii Kwa floodplain reach across three time steps.", fig.height = 4, fig.width = 10}
stacked <- terra::rast(classified)
names(stacked) <- names(classified)
terra::plot(stacked, axes = FALSE, mar = c(1, 1, 2, 1))
```

## Area Summary

The following table shows area by class for each year and the net change
between 2017 and 2023, sorted by magnitude of change.

```{r summarize}
summary_tbl <- dft_rast_summarize(classified, source = "io-lulc", unit = "ha")
```

```{r change-table}
library(dplyr)
library(tidyr)

change <- summary_tbl |>
  dplyr::select(year, class_name, area) |>
  tidyr::pivot_wider(names_from = year, values_from = area, values_fill = list(area = 0)) |>
  dplyr::mutate(
    change = `2023` - `2017`,
    pct_change = round(change / `2017` * 100, 1)
  ) |>
  dplyr::arrange(dplyr::desc(abs(change)))

knitr::kable(change, digits = 2, caption = "Net land cover change 2017--2023 (ha), sorted by absolute change.")
```

## Vegetation Change

Trees and Rangeland show the clearest signal below — tree cover
declining while rangeland expands.

```{r plot-vegetation, fig.cap = "Dominant vegetation classes over time in the Neexdzii Kwa floodplain.", fig.height = 4, fig.width = 8}
library(ggplot2)

summary_tbl |>
  dplyr::filter(class_name %in% c("Trees", "Rangeland")) |>
  ggplot(aes(x = year, y = area, fill = year)) +
  geom_col() +
  facet_wrap(~class_name, scales = "free_y") +
  scale_fill_brewer(palette = "YlGnBu") +
  labs(y = "Area (ha)", x = NULL, fill = "Year",
       title = "Vegetation cover in Neexdzii Kwa floodplain") +
  theme_minimal()
```

## Transition Detection

`dft_rast_transition()` compares two rasters cell-by-cell and returns a
transition raster plus a summary table. The first table below shows the
area that remained in the same class (stable pixels), while the second
shows pixels that changed class. Only transitions representing more than
1% of the total area are shown.

```{r transition-compute}
result <- dft_rast_transition(classified, from = "2017", to = "2023")

stable <- result$summary |>
  dplyr::filter(from_class == to_class) |>
  dplyr::filter(pct >= 1) |>
  dplyr::arrange(dplyr::desc(area))

changed <- result$summary |>
  dplyr::filter(from_class != to_class) |>
  dplyr::filter(pct >= 1) |>
  dplyr::arrange(dplyr::desc(area))
```

```{r transition-stable}
knitr::kable(stable, digits = 2,
             caption = "Stable land cover 2017--2023 (only transitions >1% of total area shown).")
```

```{r transition-change}
knitr::kable(changed, digits = 2,
             caption = "Land cover transitions 2017--2023 (only transitions >1% of total area shown).")
```

### Grouping Classes for Domain-Specific Analysis

Fine-grained LULC classes can be grouped into categories relevant to
a specific analysis. Here we demonstrate grouping Crops, Rangeland,
and Bare Ground as "Agriculture" — at 10 m resolution these classes
can represent different phases of the same land use depending on
satellite overpass timing.

The table below shows the area of Trees in 2017 that transitioned to
agriculture-related classes by 2023.

```{r tree-loss}
ag_classes <- c("Crops", "Rangeland", "Bare Ground")

# All transitions from Trees to get total Trees-origin pixel count
all_from_trees <- dft_rast_transition(classified, from = "2017", to = "2023",
                                       from_class = "Trees")
total_tree_cells <- sum(all_from_trees$summary$n_cells)

# Filter to agriculture classes
tree_loss <- dft_rast_transition(classified, from = "2017", to = "2023",
                                  from_class = "Trees",
                                  to_class = ag_classes)

# Relabel as Agriculture and compute pct of all Trees-origin pixels
tree_loss_tbl <- tree_loss$summary |>
  dplyr::mutate(to_class = "Agriculture") |>
  dplyr::group_by(from_class, to_class) |>
  dplyr::summarize(n_cells = sum(n_cells), area = sum(area), .groups = "drop") |>
  dplyr::mutate(pct_of_trees = round(n_cells / total_tree_cells * 100, 2))

knitr::kable(tree_loss_tbl, digits = 2,
             caption = "Tree loss to agriculture (Crops + Rangeland + Bare Ground) 2017--2023. Percent is of all pixels classified as Trees in 2017.")
```

### Transition Raster

The figure below maps pixels that changed class between 2017 and 2023.
Only transitions representing more than 1% of the total area are shown;
minor transitions are masked. The AOI outline is shown in red.

```{r plot-transition, fig.cap = "Spatial distribution of land cover transitions 2017--2023 (only transitions >1% of total area shown).", fig.height = 4, fig.width = 7}
trans_vals <- terra::values(result$raster)[, 1]
lvls <- terra::cats(result$raster)[[1]]

# Get codes for transitions >= 1% (excluding stable)
sig_labels <- changed$from_class  # already filtered to >1% and from != to
sig_transitions <- paste0(changed$from_class, " -> ", changed$to_class)
sig_codes <- lvls$id[lvls$transition %in% sig_transitions]

# Mask everything except significant transitions
change_vals <- rep(NA_integer_, length(trans_vals))
change_vals[trans_vals %in% sig_codes] <- trans_vals[trans_vals %in% sig_codes]
r_change <- terra::rast(result$raster)
terra::values(r_change) <- change_vals

# Keep only significant factor levels
change_lvls <- lvls[lvls$id %in% sig_codes, , drop = FALSE]
terra::set.cats(r_change, layer = 1, value = change_lvls)

terra::plot(r_change, main = "Land cover transitions 2017\u20132023",
            axes = FALSE, mar = c(1, 1, 2, 6))
plot(sf::st_geometry(sf::st_transform(aoi, terra::crs(r_change))),
     add = TRUE, border = "red", lwd = 2)
```

## Filtering Classification Noise

At 10 m resolution, many detected transitions are single-pixel or
small-cluster noise from field-forest edge effects, seasonal canopy
variation, or sensor timing differences. The `patch_area_min` parameter
removes connected patches of changed pixels smaller than a threshold
(in m²) before computing the summary.

```{r patch-filter-compare}
patch_min <- 5000
n_pixels <- patch_min / prod(terra::res(classified[[1]]))

result_filtered <- dft_rast_transition(classified, from = "2017", to = "2023",
                                       patch_area_min = patch_min)

changed_filtered <- result_filtered$summary |>
  dplyr::filter(from_class != to_class) |>
  dplyr::filter(pct >= 1) |>
  dplyr::arrange(dplyr::desc(area))

# Comparison table: unfiltered vs filtered
comparison <- changed |>
  dplyr::select(from_class, to_class, n_cells, area) |>
  dplyr::left_join(
    changed_filtered |>
      dplyr::select(from_class, to_class,
                    n_cells_filtered = n_cells, area_filtered = area),
    by = c("from_class", "to_class")
  ) |>
  dplyr::mutate(
    dplyr::across(c(n_cells_filtered, area_filtered), ~tidyr::replace_na(.x, 0)),
    cells_removed = n_cells - n_cells_filtered,
    area_removed = area - area_filtered
  )
```

Filtering at `r format(patch_min, big.mark = ",")` m²
(`r as.integer(n_pixels)` pixels at 10 m resolution) removed
`r sum(comparison$cells_removed)` pixels
(`r round(sum(comparison$area_removed), 2)` ha) of small isolated
changes. The table below compares unfiltered and filtered results.

```{r transition-filtered}
knitr::kable(comparison, digits = 2, col.names = c(
  "From", "To", "Cells", "Area (ha)", "Cells (filtered)",
  "Area (filtered)", "Cells removed", "Area removed"
), caption = paste0(
  "Land cover transitions 2017--2023: unfiltered vs filtered (min patch area ",
  format(patch_min, big.mark = ","), " m\u00b2)."))
```

The figure below shows three views: unfiltered transitions, what the
filter removed (`$removed`), and the filtered result. The `$removed`
raster is returned directly by `dft_rast_transition()` when
`patch_area_min` is set.

```{r plot-patch-filter, fig.cap = paste0("Transition raster before and after minimum patch area filtering (", format(patch_min, big.mark = ","), " m\u00b2). Centre panel shows removed patches."), fig.height = 4, fig.width = 14}
aoi_proj <- sf::st_geometry(sf::st_transform(aoi, terra::crs(r_change)))

par(mfrow = c(1, 3))
terra::plot(r_change, main = "Unfiltered", axes = FALSE, mar = c(1, 1, 2, 6))
plot(aoi_proj, add = TRUE, border = "red", lwd = 2)
terra::plot(result_filtered$removed, main = "Removed patches", axes = FALSE,
            mar = c(1, 1, 2, 6))
plot(aoi_proj, add = TRUE, border = "red", lwd = 2)
terra::plot(result_filtered$raster, main = paste0("Filtered (min ",
            format(patch_min, big.mark = ","), " m\u00b2)"),
            axes = FALSE, mar = c(1, 1, 2, 6))
plot(aoi_proj, add = TRUE, border = "red", lwd = 2)
```

## Vector Patches

`dft_transition_vectors()` converts the transition raster into `sf`
polygons — one row per connected patch. This is the format needed for
GIS QA (click patches, filter by size) and spatial attribution to
management zones.

```{r transition-vectors}
patches <- dft_transition_vectors(result$raster)

# Only actual changes (exclude same-class "transitions")
patches_changed <- patches[grepl("->", patches$transition) &
  !sapply(strsplit(patches$transition, " -> "), \(x) x[1] == x[2]), ]

knitr::kable(
  head(sf::st_drop_geometry(patches_changed[order(-patches_changed$area_ha), ]), 10),
  digits = 2,
  caption = "Ten largest change patches (same-class transitions excluded)."
)
```

When `zones` is supplied, each patch is intersected with the zone
polygons. Here we use the floodplain AOI as a single zone — in
practice this would be sub-basins, parcels, or management units.

```{r transition-zones}
aoi$zone <- "Neexdzii Kwa floodplain"
patches_zoned <- dft_transition_vectors(result$raster, zones = aoi,
                                        zone_col = "zone")
cat("Patches inside AOI:", nrow(patches_zoned), "of", nrow(patches), "total\n")
```

## Geometric Artifacts

`patch_area_min` is a one-dimensional test on a two-dimensional problem.
Area is width times length, and the classification artifacts that inflate
change on a floodplain are pathological in *width*: a class boundary that
moved by one pixel between epochs leaves a one-pixel band of "change" along
its whole length, and a channel that shifted leaves `Trees -> Water` on one
bank and `Water -> Trees` on the other. A one-pixel band along a 5 km bank
at 30 m is 15 ha — thirty times the 5,000 m² threshold used above — while a
genuine 0.5 ha cutblock is discarded as speckle. No value of
`patch_area_min` fixes both.

`dft_transition_artifact()` adds evidence for three geometric signatures to
the patches from `dft_transition_vectors()` and drops nothing — the caller
decides. All three are read from the transition raster alone, so they cost
no extra fetch.

```{r artifact-tag}
changes <- dft_transition_vectors(result$raster, changes_only = TRUE)
tagged <- dft_transition_artifact(changes, result$raster)

by_sliver <- sf::st_drop_geometry(tagged) |>
  dplyr::mutate(width = ifelse(flag_sliver, "Under 1.5 px", "1.5 px or wider")) |>
  dplyr::group_by(width) |>
  dplyr::summarise(n_patches = dplyr::n(), area_ha = sum(area_ha), .groups = "drop") |>
  dplyr::mutate(pct_patches = 100 * n_patches / sum(n_patches),
                pct_area = 100 * area_ha / sum(area_ha))

knitr::kable(by_sliver, digits = 1, col.names = c(
  "Effective width", "Patches", "Area (ha)", "% of patches", "% of area"
), caption = "Change patches 2017--2023 by effective width (2 x area / perimeter).")
```

**Sliver width.** Effective width `2 * area / perimeter` is about the short
side of a thin rectangle: a single pixel scores 0.5 px, a one-pixel band just
under 1 px, a 20 x 20 block 10 px. On this reach
`r sum(tagged$flag_sliver)` of `r nrow(tagged)` change patches are under
1.5 px wide, but they hold only
`r round(100 * sum(tagged$area_ha[tagged$flag_sliver]) / sum(tagged$area_ha))`%
of the change area — many patches, little ground.

**Boundary-hugging.** Width alone cannot tell a misregistration band from a
new road or a harvested buffer strip, which are also thin. The discriminator
is *where* the thin patch sits: a registration artifact traces a boundary
that already existed in the earlier epoch, while a real thin change cuts
across one. `boundary_frac` is the share of a patch's cells adjacent (within
`boundary_dist_max` cells) to a from-epoch cell of the patch's *to* class.
Here `r sum(tagged$flag_sliver & tagged$flag_boundary)` of the
`r sum(tagged$flag_sliver)` slivers trace a pre-existing boundary; the other
`r sum(tagged$flag_sliver & !tagged$flag_boundary)` do not, and deserve a
look before being dismissed.

**Reciprocity.** An `A -> B` patch with a `B -> A` partner of comparable area
nearby is the channel-shift signature — the net change is close to zero. The
two bands sit on opposite banks with the stable channel between them, so the
search is by proximity (`reciprocal_dist_max`, in pixels) rather than
adjacency. This is a relationship between two patches, which no per-pixel
test can express. `r sum(tagged$flag_reciprocal)` patches here have a
partner.

```{r plot-artifact, fig.cap = "Change patches 2017--2023 by artifact evidence. Slivers on a pre-existing boundary (red) are the misregistration shape; slivers that cross class boundaries (orange) are thin but not explained by a shift; patches with a reciprocal partner (purple) are the channel-shift shape.", fig.height = 5, fig.width = 7}
tagged$evidence <- dplyr::case_when(
  tagged$flag_sliver & tagged$flag_reciprocal ~ "Sliver, reciprocal partner",
  tagged$flag_sliver & tagged$flag_boundary ~ "Sliver on a pre-existing boundary",
  tagged$flag_sliver ~ "Sliver, crosses classes",
  TRUE ~ "1.5 px or wider"
)
pal <- c(
  "Sliver on a pre-existing boundary" = "#d7301f",
  "Sliver, reciprocal partner" = "#7b3294",
  "Sliver, crosses classes" = "#fdae61",
  "1.5 px or wider" = "#9e9e9e"
)
par(mar = c(0, 0, 0, 0))
plot(sf::st_geometry(tagged), col = pal[tagged$evidence],
     border = pal[tagged$evidence], lwd = 2)
plot(aoi_proj, add = TRUE, border = "black", lwd = 1)
legend("topleft", fill = pal, legend = names(pal), bty = "n", cex = 0.8)
```

The two axes compose. Below are the change patches that survive the
`r format(patch_min, big.mark = ",")` m² area filter used above, with their
artifact evidence: one of them is a `Trees -> Rangeland` band 0.75 ha in
area and under 1.5 px wide that traces the 2017 forest edge — exactly the
patch the area filter exists to remove, walking through it on length alone.
With width and boundary evidence in hand, `patch_area_min` can be lowered to
recover small real changes without readmitting the slivers.

```{r artifact-survivors}
large <- dft_transition_vectors(result$raster, changes_only = TRUE,
                                patch_area_min = patch_min)
large <- dft_transition_artifact(large, result$raster)

knitr::kable(
  sf::st_drop_geometry(large)[, c("patch_id", "transition", "area_ha", "width_px",
                                  "boundary_frac", "flag_sliver", "flag_boundary")],
  digits = 2,
  col.names = c("Patch", "Transition", "Area (ha)", "Width (px)",
                "Boundary frac.", "Sliver", "On boundary"),
  caption = paste0("Change patches surviving patch_area_min = ",
                   format(patch_min, big.mark = ","), " m\u00b2, with artifact evidence.")
)
```

The independent check is spectral. The *Trajectories as a Check on
Land-Cover Change* vignette runs `dft_rast_break()` on a Sentinel-2 kNDVI
cube over the same reach and reads a mapped `Trees -> Rangeland` outline
with no spectral break as a label change rather than trees coming off. That
route confirms or refutes individual patches but needs a cube, cannot
separate riparian deciduous trees from grass in peak summer, and cannot see
the reciprocal relationship at all. The geometric tags are the free,
categorical-only complement — use them to rank what to look at, and the
trajectories to settle it.

## Interactive Map

Toggle between classified time periods and overlay tree loss transition
layers to ground-truth change against multiple satellite basemaps.

```{r map-interactive, fig.cap = "Classified land cover by year (radio toggle) with tree loss transitions overlaid as toggleable layers. Use the fullscreen button (top left) to expand the map and access transition toggles in the layer control (top right).", fig.height = 5}
tree_trans <- dft_rast_transition(classified, from = "2017", to = "2023",
                                  from_class = "Trees")
dft_map_interactive(classified, aoi = aoi, transition = tree_trans,
                    legend_position = "bottomleft")
```
