---
title: "Land Cover Change Detection for Floodplains"
output: bookdown::html_document2
vignette: >
  %\VignetteIndexEntry{Land Cover Change Detection for Floodplains}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the drift pipeline using a small floodplain reach
on Neexdzii Kwa (Upper Bulkley River) in northern BC. We compare Esri IO LULC
land cover across 2017, 2020, and 2023 to track vegetation and land use change
in the riparian zone.

The AOI polygon was delineated using the
[flooded](https://github.com/NewGraphEnvironment/flooded) package, which
identifies floodplain extents from DEMs and stream networks.

The example data ships with the package — no STAC queries or database
connections needed.

## Load Data

```{r load-data}
library(drift)
library(terra)
library(sf)

# AOI polygon (floodplain delineated via flooded package)
aoi <- sf::st_read(
  system.file("extdata", "example_aoi.gpkg", package = "drift"),
  quiet = TRUE
)

# IO LULC rasters for 3 years
years <- c(2017, 2020, 2023)
rasters <- lapply(years, function(yr) {
  terra::rast(system.file("extdata", paste0("example_", yr, ".tif"),
                          package = "drift"))
})
names(rasters) <- years
```

## Classify

Apply IO LULC class names and colors from the shipped class table.

```{r classify}
classified <- dft_rast_classify(rasters, source = "io-lulc")

# Check factor levels
terra::levels(classified[["2020"]])[[1]]
```

## Classified Rasters

The figure below shows the three classified time steps side by side.

```{r plot-classified, fig.cap = "Classified land cover for the Neexdzii Kwa floodplain reach across three time steps.", fig.height = 4, fig.width = 10}
stacked <- terra::rast(classified)
names(stacked) <- names(classified)
terra::plot(stacked, axes = FALSE, mar = c(1, 1, 2, 1))
```

## Area Summary

The following table shows area by class for each year and the net change
between 2017 and 2023, sorted by magnitude of change.

```{r summarize}
summary_tbl <- dft_rast_summarize(classified, source = "io-lulc", unit = "ha")
```

```{r change-table}
library(dplyr)
library(tidyr)

change <- summary_tbl |>
  dplyr::select(year, class_name, area) |>
  tidyr::pivot_wider(names_from = year, values_from = area, values_fill = list(area = 0)) |>
  dplyr::mutate(
    change = `2023` - `2017`,
    pct_change = round(change / `2017` * 100, 1)
  ) |>
  dplyr::arrange(dplyr::desc(abs(change)))

knitr::kable(change, digits = 2, caption = "Net land cover change 2017--2023 (ha), sorted by absolute change.")
```

## Vegetation Change

Trees and Rangeland show the clearest signal below — tree cover
declining while rangeland expands.

```{r plot-vegetation, fig.cap = "Dominant vegetation classes over time in the Neexdzii Kwa floodplain.", fig.height = 4, fig.width = 8}
library(ggplot2)

summary_tbl |>
  dplyr::filter(class_name %in% c("Trees", "Rangeland")) |>
  ggplot(aes(x = year, y = area, fill = year)) +
  geom_col() +
  facet_wrap(~class_name, scales = "free_y") +
  scale_fill_brewer(palette = "YlGnBu") +
  labs(y = "Area (ha)", x = NULL, fill = "Year",
       title = "Vegetation cover in Neexdzii Kwa floodplain") +
  theme_minimal()
```

## Transition Detection

`dft_rast_transition()` compares two rasters cell-by-cell and returns a
transition raster plus a summary table. The first table below shows the
area that remained in the same class (stable pixels), while the second
shows pixels that changed class. Only transitions representing more than
1% of the total area are shown.

```{r transition-compute}
result <- dft_rast_transition(classified, from = "2017", to = "2023")

stable <- result$summary |>
  dplyr::filter(from_class == to_class) |>
  dplyr::filter(pct >= 1) |>
  dplyr::arrange(dplyr::desc(area))

changed <- result$summary |>
  dplyr::filter(from_class != to_class) |>
  dplyr::filter(pct >= 1) |>
  dplyr::arrange(dplyr::desc(area))
```

```{r transition-stable}
knitr::kable(stable, digits = 2,
             caption = "Stable land cover 2017--2023 (only transitions >1% of total area shown).")
```

```{r transition-change}
knitr::kable(changed, digits = 2,
             caption = "Land cover transitions 2017--2023 (only transitions >1% of total area shown).")
```

### Grouping Classes for Domain-Specific Analysis

Fine-grained LULC classes can be grouped into categories relevant to
a specific analysis. Here we demonstrate grouping Crops, Rangeland,
and Bare Ground as "Agriculture" — at 10 m resolution these classes
can represent different phases of the same land use depending on
satellite overpass timing.

The table below shows the area of Trees in 2017 that transitioned to
agriculture-related classes by 2023.

```{r tree-loss}
ag_classes <- c("Crops", "Rangeland", "Bare Ground")

# All transitions from Trees to get total Trees-origin pixel count
all_from_trees <- dft_rast_transition(classified, from = "2017", to = "2023",
                                       from_class = "Trees")
total_tree_cells <- sum(all_from_trees$summary$n_cells)

# Filter to agriculture classes
tree_loss <- dft_rast_transition(classified, from = "2017", to = "2023",
                                  from_class = "Trees",
                                  to_class = ag_classes)

# Relabel as Agriculture and compute pct of all Trees-origin pixels
tree_loss_tbl <- tree_loss$summary |>
  dplyr::mutate(to_class = "Agriculture") |>
  dplyr::group_by(from_class, to_class) |>
  dplyr::summarize(n_cells = sum(n_cells), area = sum(area), .groups = "drop") |>
  dplyr::mutate(pct_of_trees = round(n_cells / total_tree_cells * 100, 2))

knitr::kable(tree_loss_tbl, digits = 2,
             caption = "Tree loss to agriculture (Crops + Rangeland + Bare Ground) 2017--2023. Percent is of all pixels classified as Trees in 2017.")
```

### Transition Raster

The figure below maps pixels that changed class between 2017 and 2023.
Only transitions representing more than 1% of the total area are shown;
minor transitions are masked. The AOI outline is shown in red.

```{r plot-transition, fig.cap = "Spatial distribution of land cover transitions 2017--2023 (only transitions >1% of total area shown).", fig.height = 4, fig.width = 7}
trans_vals <- terra::values(result$raster)[, 1]
lvls <- terra::cats(result$raster)[[1]]

# Get codes for transitions >= 1% (excluding stable)
sig_labels <- changed$from_class  # already filtered to >1% and from != to
sig_transitions <- paste0(changed$from_class, " -> ", changed$to_class)
sig_codes <- lvls$id[lvls$transition %in% sig_transitions]

# Mask everything except significant transitions
change_vals <- rep(NA_integer_, length(trans_vals))
change_vals[trans_vals %in% sig_codes] <- trans_vals[trans_vals %in% sig_codes]
r_change <- terra::rast(result$raster)
terra::values(r_change) <- change_vals

# Keep only significant factor levels
change_lvls <- lvls[lvls$id %in% sig_codes, , drop = FALSE]
terra::set.cats(r_change, layer = 1, value = change_lvls)

terra::plot(r_change, main = "Land cover transitions 2017\u20132023",
            axes = FALSE, mar = c(1, 1, 2, 6))
plot(sf::st_geometry(sf::st_transform(aoi, terra::crs(r_change))),
     add = TRUE, border = "red", lwd = 2)
```

## Filtering Classification Noise

At 10 m resolution, many detected transitions are single-pixel or
small-cluster noise from field-forest edge effects, seasonal canopy
variation, or sensor timing differences. The `patch_area_min` parameter
removes connected patches of changed pixels smaller than a threshold
(in m²) before computing the summary.

```{r patch-filter-compare}
patch_min <- 5000
n_pixels <- patch_min / prod(terra::res(classified[[1]]))

result_filtered <- dft_rast_transition(classified, from = "2017", to = "2023",
                                       patch_area_min = patch_min)

changed_filtered <- result_filtered$summary |>
  dplyr::filter(from_class != to_class) |>
  dplyr::filter(pct >= 1) |>
  dplyr::arrange(dplyr::desc(area))

# Comparison table: unfiltered vs filtered
comparison <- changed |>
  dplyr::select(from_class, to_class, n_cells, area) |>
  dplyr::left_join(
    changed_filtered |>
      dplyr::select(from_class, to_class,
                    n_cells_filtered = n_cells, area_filtered = area),
    by = c("from_class", "to_class")
  ) |>
  dplyr::mutate(
    dplyr::across(c(n_cells_filtered, area_filtered), ~tidyr::replace_na(.x, 0)),
    cells_removed = n_cells - n_cells_filtered,
    area_removed = area - area_filtered
  )
```

Filtering at `r format(patch_min, big.mark = ",")` m²
(`r as.integer(n_pixels)` pixels at 10 m resolution) removed
`r sum(comparison$cells_removed)` pixels
(`r round(sum(comparison$area_removed), 2)` ha) of small isolated
changes. The table below compares unfiltered and filtered results.

```{r transition-filtered}
knitr::kable(comparison, digits = 2, col.names = c(
  "From", "To", "Cells", "Area (ha)", "Cells (filtered)",
  "Area (filtered)", "Cells removed", "Area removed"
), caption = paste0(
  "Land cover transitions 2017--2023: unfiltered vs filtered (min patch area ",
  format(patch_min, big.mark = ","), " m\u00b2)."))
```

The figure below shows three views: unfiltered transitions, what the
filter removed (`$removed`), and the filtered result. The `$removed`
raster is returned directly by `dft_rast_transition()` when
`patch_area_min` is set.

```{r plot-patch-filter, fig.cap = paste0("Transition raster before and after minimum patch area filtering (", format(patch_min, big.mark = ","), " m\u00b2). Centre panel shows removed patches."), fig.height = 4, fig.width = 14}
aoi_proj <- sf::st_geometry(sf::st_transform(aoi, terra::crs(r_change)))

par(mfrow = c(1, 3))
terra::plot(r_change, main = "Unfiltered", axes = FALSE, mar = c(1, 1, 2, 6))
plot(aoi_proj, add = TRUE, border = "red", lwd = 2)
terra::plot(result_filtered$removed, main = "Removed patches", axes = FALSE,
            mar = c(1, 1, 2, 6))
plot(aoi_proj, add = TRUE, border = "red", lwd = 2)
terra::plot(result_filtered$raster, main = paste0("Filtered (min ",
            format(patch_min, big.mark = ","), " m\u00b2)"),
            axes = FALSE, mar = c(1, 1, 2, 6))
plot(aoi_proj, add = TRUE, border = "red", lwd = 2)
```

## Multi-Year Classification and Transition Overlay

Toggle between classified time periods and overlay tree loss transition
layers to ground-truth change against multiple satellite basemaps.

```{r map-interactive, fig.cap = "Classified land cover by year (radio toggle) with tree loss transitions overlaid as toggleable layers.", fig.height = 5}
tree_trans <- dft_rast_transition(classified, from = "2017", to = "2023",
                                  from_class = "Trees")
dft_map_interactive(classified, aoi = aoi, transition = tree_trans)
```
