---
title: "Land Cover Change — Neexdzii Kwa"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Land Cover Change — Neexdzii Kwa}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

This vignette demonstrates the drift pipeline using a small floodplain reach
on Neexdzii Kwa (Upper Bulkley River) in northern BC. We compare Esri IO LULC
land cover across 2017, 2020, and 2023 to track vegetation and land use change
in the riparian zone.

The AOI polygon was delineated using the
[flooded](https://github.com/NewGraphEnvironment/flooded) package, which
identifies floodplain extents from DEMs and stream networks.

The example data ships with the package — no STAC queries or database
connections needed.

## Load Data

```{r load-data}
library(drift)
library(terra)
library(sf)

# AOI polygon (floodplain delineated via flooded package)
aoi <- sf::st_read(
  system.file("extdata", "example_aoi.gpkg", package = "drift"),
  quiet = TRUE
)

# IO LULC rasters for 3 years
years <- c(2017, 2020, 2023)
rasters <- lapply(years, function(yr) {
  terra::rast(system.file("extdata", paste0("example_", yr, ".tif"),
                          package = "drift"))
})
names(rasters) <- years
```

## Classify

Apply IO LULC class names and colors from the shipped class table.

```{r classify}
classified <- dft_rast_classify(rasters, source = "io-lulc")

# Check factor levels
terra::levels(classified[["2020"]])[[1]]
```

## Plot Classified Rasters

```{r plot-classified, fig.height = 4, fig.width = 10}
# Stack into a single multi-layer SpatRaster for panel plot
stacked <- terra::rast(classified)
names(stacked) <- names(classified)
terra::plot(stacked, axes = FALSE, mar = c(1, 1, 2, 1))
```

## Summarize

Compute area by class for each year.

```{r summarize}
summary_tbl <- dft_rast_summarize(classified, source = "io-lulc", unit = "ha")
summary_tbl
```

## Area Change Table

```{r change-table}
library(dplyr)
library(tidyr)

change <- summary_tbl |>
  dplyr::select(year, class_name, area) |>
  tidyr::pivot_wider(names_from = year, values_from = area, values_fill = list(area = 0)) |>
  dplyr::mutate(
    change = `2023` - `2017`,
    pct_change = round(change / `2017` * 100, 1)
  ) |>
  dplyr::arrange(dplyr::desc(abs(change)))

knitr::kable(change, digits = 2, caption = "Land cover change 2017–2023 (ha)")
```

## Interactive Map

Toggle between time periods to see how land cover changed across the floodplain.

```{r map-interactive, fig.height = 5}
dft_map_interactive(classified, aoi = aoi)
```
