---
title: "Trajectories as a Check on Land-Cover Change"
output: bookdown::html_document2
vignette: >
  %\VignetteIndexEntry{Trajectories as a Check on Land-Cover Change}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

The goal here is floodplain land-cover change: understanding how riparian
vegetation on the Neexdzii Kwa (Upper Bulkley) reach changes over time, to inform
where it is worth keeping standing and where restoration is justified. The *Land
Cover Change Detection* vignette answers that with categorical labels — a pixel
mapped as Trees in 2020 and Rangeland in 2023 counts as a "Trees to Rangeland"
transition, and you can total the hectares.

Those categorical transitions are the starting point, not the last word.
Automated 10 m land-cover products are noisiest exactly where riparian cover is
sparse and deciduous — a pixel sitting on the Trees/Rangeland decision boundary
can flip its label year to year from classifier noise, not real change. This
vignette uses a **continuous** kNDVI trajectory from Sentinel-2 on the **same
reach and 2017–2023 window** to ask a different question of each pixel: is the
vegetation actually *declining*, and does the greenness back up the mapped loss?
Two reductions do the work — a monotonic **trend** ([dft_rast_trend()]) for
gradual degradation or recovery, and an abrupt **break** ([dft_rast_break()],
via BFAST) for dated disturbances.

## The pipeline

```{r pipeline, eval = FALSE}
library(drift)

aoi <- sf::st_read(
  system.file("extdata", "example_aoi.gpkg", package = "drift")
)

# Growing-season (June-September) monthly kNDVI, 2017-2023
cube <- dft_stac_cube(aoi, index = "kndvi",
                      datetime = "2017-01-01/2023-12-31", months = 6:9)

trend  <- dft_rast_trend(cube)                        # gradual: which way, how fast
breaks <- dft_rast_break(cube, start = c(2022, 1))    # abrupt: when it broke
```

This vignette loads a small pre-computed artifact (generated by
`data-raw/vignette_data_break.R`) so every chunk runs live, no network:

```{r load}
library(terra)

d        <- readRDS(system.file("testdata", "neexdzii_break.rds", package = "drift"))
trend    <- terra::unwrap(d$trend_r)
breaks   <- terra::unwrap(d$break_r)
treeloss <- terra::unwrap(d$treeloss)   # LULC Trees -> Rangeland/Bare (ground truth)
traj     <- d$traj
qa       <- d$qa
```

## Is the vegetation actually declining?

`dft_rast_trend()` gives a robust (Theil-Sen) slope per pixel — the kNDVI change
per year across 2017–2023 — resistant to a single off season. Negative is
gradual decline; positive is recovery.

```{r trend-map, fig.cap = "kNDVI trend, 2017-2023 (index change per year). Red = declining greenness, blue = recovering. Most of the reach is stable to recovering -- the open floodplain in particular is greening; genuine, significant decline is the exception.", fig.height = 5}
terra::plot(trend[["trend"]], col = hcl.colors(21, "Blue-Red 3", rev = TRUE),
            range = c(-0.04, 0.04), main = "kNDVI trend per year")
```

Across the reach the trend is stable-to-slightly-positive — the floodplain
vegetation as a whole is holding or recovering over the record, not broadly
declining. That itself is a conservation-relevant result.

## Does the greenness back up the mapped forest loss?

Now the check. Take the stands the land-cover product maps as **Trees to
Rangeland/Bare** over 2020–2023 and ask what their *continuous* trajectory did —
compared with stands mapped as staying forest, and with the rest of the reach.

```{r qa}
qa$median_trend   <- round(qa$median_trend, 4)
qa$pct_declining  <- sprintf("%.0f%%", 100 * qa$pct_declining)
qa$pct_recovering <- sprintf("%.0f%%", 100 * qa$pct_recovering)
qa[c("group", "n", "median_trend", "pct_declining", "pct_recovering")]
```

The stands the categorical product flags as lost show a **median trend of about
zero**. A small subset — roughly 1 in 20 — does carry a *significant* declining
trend, a few times the background rate, so there is some genuine loss in there.
But the great majority show **no real decline in vegetation**: the mapped "forest
loss" is dominated by borderline pixels flipping label, not canopy coming off,
and the floodplain background is in fact *greening*. The grouped trajectories make
the point — the tree-loss stands track the rest of the reach rather than diverging
downward.

```{r trajectory, fig.cap = "Mean growing-season kNDVI for LULC tree-loss stands, intact forest, and the reach background. The tree-loss stands do not fall away from the others -- the categorical transition is not matched by a loss of greenness.", fig.height = 4.2}
g <- split(traj, traj$group)
cols <- c("tree-loss (LULC)" = "firebrick", "intact forest" = "forestgreen",
          "background" = "grey55")
plot(g[[1]]$date, g[[1]]$kndvi, type = "n", ylim = c(0, 0.75),
     xlab = "", ylab = "kNDVI", main = "Growing-season kNDVI by group")
for (nm in names(cols)) {
  lines(g[[nm]]$date, g[[nm]]$kndvi, type = "o", pch = 19, cex = 0.5, col = cols[nm])
}
legend("bottomleft", names(cols), col = cols, pch = 19, lty = 1, bty = "n")
```

So the trajectory acts as a **credibility check** on the categorical area
numbers: before reporting hectares of floodplain forest lost, confirm which
transitions carry a real spectral decline. Here, most do not — a caution worth
carrying into the reporting, and a stronger evidence base for "this reach is
holding, leave it standing."

## When change is abrupt

For a genuine disturbance — a scour, fire, or clear canopy removal — the change
is a step, not a slope, and [dft_rast_break()] dates it. The break map below
outlines the mapped tree loss; where a real break coincides with it, that is a
confirmed, dated loss; where it does not, the categorical flip stands unsupported.

```{r break-map, fig.cap = "Break magnitude (red = a larger index drop) with LULC tree-loss outlined. Genuine dated breaks are sparse here -- consistent with the flat trends.", fig.height = 5}
terra::plot(breaks[["break_mag"]], col = hcl.colors(21, "Blue-Red 3", rev = TRUE),
            range = c(-0.45, 0.45), main = "kNDVI break magnitude")
terra::plot(terra::as.polygons(terra::ifel(treeloss == 1, 1, NA)),
            add = TRUE, border = "black", lwd = 1)
```

## When to use which

- **Categorical** (`dft_stac_fetch()` + `dft_rast_transition()`) — for named
  transitions and hectares: "Trees to Rangeland." The right tool for area
  accounting, but treat its transitions as hypotheses on this landscape.
- **Trend** (`dft_rast_trend()`) — for gradual degradation and recovery the
  annual labels cannot show, and as a QA layer on those labels. Directly answers
  "is this stand declining, holding, or recovering?"
- **Break** (`dft_rast_break()`) — for dating an abrupt disturbance and ranking
  severity.

A caveat specific to this reach: in peak summer, deciduous riparian trees and
grass are both green, so kNDVI cannot cleanly separate them — the continuous
tools are strongest on evergreen canopy and clear removal (conifer clearcut,
fire, scour) and on gradual trends, and weakest at telling riparian tree from
rangeland. Used together, land cover says *what*, and the trajectory says whether
it is *real*, which *way* it is going, and *when*.
