---
title: "Continuous Index-Trajectory Change Detection"
output: bookdown::html_document2
vignette: >
  %\VignetteIndexEntry{Continuous Index-Trajectory Change Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

Categorical land-cover differencing (see the *Land Cover Change Detection*
vignette) compares two classified labels per pixel. For floodplain reaches that
is the weakest available signal: an annual 9-class product collapses gravel
bars, wet/dry channel, and sparse seasonal riparian into a single class, and
annual compositing smears the timing of events.

This vignette shows the stronger alternative in drift: a **continuous index
trajectory** per pixel. Instead of comparing two labels, we build a monthly
kernel-NDVI (kNDVI) time series from Sentinel-2 and ask, per pixel, *when* the
spectral history broke and *by how much* — using
[BFAST](https://bfast2.github.io/) monitoring.

## The pipeline

The full pipeline is two calls. `dft_stac_cube()` builds a cloud-masked kNDVI
cube from the Planetary Computer, and `dft_rast_break()` reduces it over time
with `bfast::bfastmonitor()`:

```{r pipeline, eval = FALSE}
library(drift)

aoi <- sf::st_read(
  system.file("extdata", "example_aoi.gpkg", package = "drift")
)

# Growing-season (June-September) monthly kNDVI, 2018-2023
cube <- dft_stac_cube(
  aoi,
  source   = "sentinel-2-l2a",
  index    = "kndvi",
  datetime = "2018-01-01/2023-12-31",
  dt       = "P1M",
  months   = 6:9          # drop snow / low-sun winter noise, fetch ~3x fewer scenes
)

# Per-pixel breakpoint: monitoring period starts 2022, low harmonic order for
# the growing-season-only series
breaks <- dft_rast_break(cube, start = c(2022, 1), order = 1)
```

**Why the growing season?** At this latitude (~54 degrees N), winter Sentinel-2
scenes are snow-covered and low-sun -- they carry no vegetation signal and would
otherwise dominate the seasonal cycle BFAST has to fit. Restricting to June to
September (`months = 6:9`) keeps the observations where a riparian scour or
vegetation loss actually shows up, and streams roughly a third of the scenes.
The winter months become `NA` in the monthly cube, so the per-pixel series stays
regular at `frequency = 12` and BFAST still models the within-season trajectory.

That fetch requires network access and the `gdalcubes` and `bfast` packages, so
this vignette instead loads a small pre-computed artifact (generated by
`data-raw/vignette_data_break.R`) and works from it. Every chunk below runs
live, with no network.

```{r load}
library(terra)

d      <- readRDS(system.file("testdata", "neexdzii_break.rds", package = "drift"))
breaks <- terra::unwrap(d$breaks)
traj   <- d$traj    # a few representative pixel trajectories for the last panel
aoi    <- d$aoi

breaks
```

The result is a two-band raster: `break_date` (the decimal year of the detected
break, or `NA` where none was found) and `break_mag` (the signed change in
kNDVI — negative where the index dropped).

## Where and when did trajectories break?

The magnitude map is the headline. Negative values (channel scour, vegetation
loss) are the floodplain disturbances we care about; positive values are
re-establishment.

```{r mag-map, fig.cap = "kNDVI break magnitude across the Neexdzii Kwa floodplain reach. Negative (red) marks index drops -- scour or vegetation loss; positive (blue) marks establishment.", fig.height = 5}
mag <- breaks[["break_mag"]]
rng <- max(abs(terra::minmax(mag)), na.rm = TRUE)

terra::plot(
  mag, col = hcl.colors(21, "Blue-Red 3"),
  range = c(-rng, rng),
  main = "kNDVI break magnitude (negative = scour / loss)"
)
plot(sf::st_geometry(sf::st_transform(aoi, terra::crs(mag))),
     add = TRUE, border = "grey20", lwd = 1)
```

The companion date map shows *when* each break occurred — useful for tying a
disturbance to a specific freshet or storm year.

```{r date-map, fig.cap = "Decimal year of the detected kNDVI break. Pixels with no detected break are transparent.", fig.height = 5}
terra::plot(breaks[["break_date"]], col = hcl.colors(20, "Viridis"),
            main = "Break year")
plot(sf::st_geometry(sf::st_transform(aoi, terra::crs(breaks))),
     add = TRUE, border = "grey20", lwd = 1)
```

## Reading a single pixel's trajectory

The break map is a per-pixel reduction of a full monthly time series. To see
what BFAST is responding to, the artifact ships two representative pixel
trajectories: the pixel with the strongest detected drop, and a stable pixel
where no break was found. The dashed line marks the detected break date on the
scour pixel.

```{r trajectory, fig.cap = "Monthly kNDVI trajectories for the strongest-drop pixel (scour) and a stable pixel. The dashed line marks the detected break; the sustained drop after it is what drives the negative magnitude.", fig.height = 4}
scour  <- traj[traj$label == "scour", ]
stable <- traj[traj$label == "stable", ]

plot(scour$date, scour$kndvi, type = "o", pch = 19, cex = 0.5, col = "firebrick",
     ylim = range(traj$kndvi, na.rm = TRUE), xlab = "", ylab = "kNDVI",
     main = "kNDVI trajectories: scour vs stable pixel")
lines(stable$date, stable$kndvi, type = "o", pch = 19, cex = 0.5, col = "grey40")

break_yr <- scour$break_date[1]
if (is.finite(break_yr)) {
  abline(v = as.Date(paste0(floor(break_yr), "-01-01")) +
           (break_yr - floor(break_yr)) * 365, lty = 2, col = "firebrick")
}
legend("bottomleft", c("scour pixel", "stable pixel"), bty = "n",
       col = c("firebrick", "grey40"), pch = 19, lty = 1)
```

## When to use which

Reach for `dft_rast_break()` over categorical differencing when timing and
severity matter — dating a scour event, ranking reaches by disturbance
magnitude, or catching change that annual class labels miss. Reach for the
categorical pipeline when you need named land-cover transitions (e.g. trees to
rangeland) for reporting. The two are complementary: the optional bolt-on of
labelling breaks with from/to classes bridges them.
