---
title: "Continuous Index-Trajectory Change Detection"
output: bookdown::html_document2
vignette: >
  %\VignetteIndexEntry{Continuous Index-Trajectory Change Detection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

The *Land Cover Change Detection* vignette tracks change on the Neexdzii Kwa
(Upper Bulkley) floodplain by comparing categorical land-cover labels across
annual snapshots (2017, 2020, 2023). That answers *what* changed and *roughly
when* — a stand mapped as Trees in 2020 and Rangeland in 2023 was cleared
*somewhere* in that three-year gap.

This vignette runs the **continuous** alternative on the **same reach and the
same 2017–2023 window**, so the two are directly comparable. Instead of two
categorical labels, it builds a monthly kernel-NDVI (kNDVI) trajectory per pixel
from Sentinel-2 and asks, with [BFAST](https://bfast2.github.io/) monitoring,
*when* each pixel's spectral history broke and by how much. The payoff over the
annual product is **timing**: it dates the change to a season, not a three-year
bracket.

## The pipeline

```{r pipeline, eval = FALSE}
library(drift)

aoi <- sf::st_read(
  system.file("extdata", "example_aoi.gpkg", package = "drift")
)

# Growing-season (June-September) monthly kNDVI, 2017-2023
cube <- dft_stac_cube(
  aoi,
  source   = "sentinel-2-l2a",
  index    = "kndvi",
  datetime = "2017-01-01/2023-12-31",
  dt       = "P1M",
  months   = 6:9          # drop snow / low-sun winter noise, ~3x fewer scenes
)

# Per-pixel break: monitor from 2022 with 2017-2021 as the stable history
breaks <- dft_rast_break(cube, start = c(2022, 1), order = 1)
```

`dft_stac_cube()` restricts to the growing season (winter Sentinel-2 at this
latitude is snow and low-sun noise) and, because the Sentinel-2 reflectance
offset changed at processing baseline 04.00 (2022-01-25), splits the scenes at
that boundary and corrects each side — so the 2017–2021 history and the 2022+
monitoring period sit on one consistent scale with no artificial step.

Because that fetch needs the network plus `gdalcubes` and `bfast`, this vignette
loads a small pre-computed artifact (generated by `data-raw/vignette_data_break.R`)
and works from it — every chunk below runs live, no network.

```{r load}
library(terra)

d        <- readRDS(system.file("testdata", "neexdzii_break.rds", package = "drift"))
breaks   <- terra::unwrap(d$breaks)
treeloss <- terra::unwrap(d$treeloss)   # LULC Trees -> Rangeland/Bare (ground truth)
traj     <- d$traj
st       <- d$stats
aoi      <- d$aoi

breaks
```

## When and how much did trajectories break?

`break_mag` is the headline: negative where kNDVI dropped (vegetation loss),
positive where it rose. `break_date` is the decimal year of the detected break.

```{r mag-map, fig.cap = "kNDVI break magnitude across the reach (negative = index drop). Black outlines are stands the land-cover product maps as Trees to Rangeland/Bare over 2017-2023 -- the tree loss falls preferentially on negative breaks.", fig.height = 5}
mag <- breaks[["break_mag"]]
terra::plot(mag, col = hcl.colors(21, "Blue-Red 3"), range = c(-0.45, 0.45),
            main = "kNDVI break magnitude")
terra::plot(terra::as.polygons(terra::ifel(treeloss == 1, 1, NA)),
            add = TRUE, border = "black", lwd = 1)
```

```{r date-map, fig.cap = "Decimal year of the detected break. Most breaks -- and most of the tree loss -- date to the 2023 growing season.", fig.height = 5}
terra::plot(breaks[["break_date"]], col = hcl.colors(20, "Viridis"),
            main = "Break year")
```

## Does it agree with the land-cover product?

We can check the continuous result against the categorical one directly: take the
stands the land-cover product maps as **Trees to Rangeland/Bare** (2020 to 2023)
and ask what BFAST did there.

```{r agreement}
data.frame(
  metric = c("break rate", "median break magnitude", "median break date"),
  tree_loss = c(sprintf("%.0f%%", 100 * st$rate_treeloss),
                sprintf("%.3f", st$mag_treeloss),
                sprintf("%.1f", st$date_treeloss)),
  background = c(sprintf("%.0f%%", 100 * st$rate_background),
                 sprintf("%.3f", st$mag_background), "--")
)
```

The mapped tree-loss stands break about **twice as often** and with a
**stronger** negative magnitude than the background, and BFAST dates that loss to
the **2023 growing season** — timing the annual product structurally cannot
provide. The grouped trajectories show it: tree-loss pixels (red) drop below both
the intact forest (green) and the reach as a whole in 2023.

```{r trajectory, fig.cap = "Mean growing-season kNDVI for LULC tree-loss stands, intact forest, and the reach background. Intact forest stays high; tree-loss stands diverge downward in 2023. A basin-wide dip that year (a dry / wildfire-smoke season) sits under everything -- the categorical product cannot separate that from stand-level loss, but the trajectory shows both.", fig.height = 4.2}
g <- split(traj, traj$group)
cols <- c("tree-loss (LULC)" = "firebrick", "intact forest" = "forestgreen",
          "background" = "grey55")
plot(g[[1]]$date, g[[1]]$kndvi, type = "n", ylim = c(0, 0.75),
     xlab = "", ylab = "kNDVI", main = "Growing-season kNDVI by group")
for (nm in names(cols)) {
  lines(g[[nm]]$date, g[[nm]]$kndvi, type = "o", pch = 19, cex = 0.5, col = cols[nm])
}
abline(v = as.Date("2022-01-01"), lty = 3)
legend("bottomleft", names(cols), col = cols, pch = 19, lty = 1, bty = "n")
```

## When to use which

The two pipelines are complements, not competitors. Reach for the **categorical**
path (`dft_stac_fetch()` + `dft_rast_transition()`) when you need named
land-cover transitions for reporting — "Trees to Rangeland" is legible and maps
cleanly. Reach for the **trajectory** path here when timing and severity matter:
it dates a disturbance to the season, ranks stands by magnitude, and — as the
2023 dip shows — surfaces gradual, sub-annual signals the annual labels miss.
Read together, land cover says *what*, and the trajectory says *exactly when*.
