---
title: "What a Land-Cover Change Figure Is Made Of"
output: bookdown::html_document2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  dpi = 150
)
```

```{r load}
library(drift)
d <- function(f) file.path(system.file("extdata", "temporal-composition", package = "drift"), f)

grp <- utils::read.csv(d("summary_groups.csv"), stringsAsFactors = FALSE)
tree <- utils::read.csv(d("summary_treeloss_temporal.csv"), stringsAsFactors = FALSE)
grid <- utils::read.csv(d("bulk_grid_1km.csv"), stringsAsFactors = FALSE)
win <- utils::read.csv(d("bulk_window.csv"), stringsAsFactors = FALSE)
art <- readRDS(d("bulk_window.rds"))

# colours come from the gq registry, so no hex literal appears in this article
reg <- gq::gq_reg_merge(gq::gq_reg_main(),
                        gq::gq_reg_custom(system.file("cartography", "drift_temporal.csv",
                                                      package = "drift")))
tcl <- gq::gq_tmap_classes(reg$layers$temporal_category)
pal <- tcl$values                      # named by class_value, in class order
lab <- stats::setNames(tcl$labels, names(pal))
# a malformed row in the registry CSV adds a phantom layer keyed by its own text
# and gq accepts it without a warning, so assert the class set rather than
# assuming the file parsed as intended
stopifnot(identical(names(pal),
                    c("stable", "break_sustained", "break_endpoint", "flicker", "stable_flicker")))

b <- grp[grp$group == "bulk", ]
# published watershed-group codes. Only BULK is named in prose, because only
# its full name is established here; inventing the other three would be a fact
# with no source.
group_name <- stats::setNames(toupper(grp$group), grp$group)
```

A published floodplain land-cover change layer is a comparison of two years. For the Bulkley
floodplain (BULK) it says that **`r format(round(b$changed_ha), big.mark = ",")` ha** — about
`r round(b$pct_changed_of_valid)`% of the mapped floodplain — carried a different class in 2023
than in 2017. That number is the input to most things anyone wants to ask.

This article is about what is inside it. It is not about *why* land cover changed: attributing a
patch to a dated fire or cutblock is a separate stage of the same production run, and its results
are written up separately. What follows is the prior question — whether the change a two-year
comparison reports is a change that stayed.

## Looking at every year, not just the ends

The published series carries a classified raster for **every** year from 2017 to 2023, not only
the two endpoints. So each pixel can be read as a seven-year sequence rather than a before-and-
after pair, and sorted by how it behaves:

- **Sustained change** — the class switches once, and holds for at least two years on each side.
- **Endpoint-only change** — the class switches once, but one endpoint year is the only year that
  differs. A real change in the final year of the series looks exactly like this, so it is not the
  same thing as an error.
- **Unsettled** — the class changes more than once and never settles.

A fourth group matters too: pixels that flicker between classes but happen to hold the *same*
class in 2017 and 2023. A two-year comparison cannot see them at all.

```{r fig-patch, fig.height = 4.4, fig.cap = paste0("One ", round(win$area_ha, 2), " ha patch of the Bulkley floodplain mapped as Trees in 2017 and Rangeland in 2023, shown for each year of the series. The outline is the patch; classes are Esri IO LULC. The patch was selected by rule, not by eye: of ", format(win$n_candidates_all, big.mark = ","), " change patches, ", win$n_candidates_transition, " are Trees to Rangeland, ", win$n_candidates_area, " of those are 1 to 4 ha, and ", win$n_candidates_balanced, " of those carry all three temporal categories at 10% or more of their cells. This is the most evenly split of the ", win$n_candidates_balanced, ".")}
lulc <- terra::unwrap(art$lulc)
ct <- dft_class_table("io-lulc")
present <- sort(unique(stats::na.omit(as.vector(terra::values(lulc)))))
ct <- ct[match(present, ct$code), ]
cols <- ct$color

# A colour TABLE, keyed by cell value. `type = "classes"` with a levels/col pair
# maps each layer's own sorted unique values onto the vector positionally, so a
# year that happens to contain no Snow/Ice shifts every class after it -- 2023's
# Rangeland drew in Snow/Ice's blue. Five of the seven years are missing a class,
# so five of seven panels were wrong, and only reading the picture showed it.
tab <- data.frame(value = ct$code, col = ct$color)
for (i in seq_len(terra::nlyr(lulc))) terra::coltab(lulc, layer = i) <- tab

op <- graphics::par(mfrow = c(2, 4), mar = c(0.4, 0.4, 1.6, 0.4), oma = c(0, 0, 0, 0))
for (i in seq_len(terra::nlyr(lulc))) {
  terra::plot(lulc[[i]], legend = FALSE, axes = FALSE, mar = NA, main = names(lulc)[i])
  plot(sf::st_geometry(art$patch), add = TRUE, border = "black", lwd = 1.6)
}
graphics::plot.new()
graphics::legend("center", legend = ct$class_name, fill = cols, bty = "n", cex = 0.95)
graphics::par(op)
```

The patch above is mapped as Trees in 2017 and Rangeland in 2023, so the
published layer counts all `r round(win$area_ha, 2)` ha of it as tree loss. Reading across the
years, only part of it behaves that way.

## The same split across a whole floodplain

```{r fig-map, fig.height = 3.6, fig.cap = paste0("Bulkley floodplain change by temporal category. Left: the whole floodplain, aggregated to 1 km cells and shaded by the share of that cell's changed area that never settles; the box marks the detail at right. Right: a 4 km reach at the mapped 10 m resolution. The floodplain is the drawn area — there is no basemap, so nothing outside the mapped extent is shown.")}
op <- graphics::par(mfrow = c(1, 2), mar = c(0.4, 0.4, 1.8, 0.4))

sub <- grid[grid$sustained_ha + grid$endpoint_ha + grid$unsettled_ha > 0, ]
frac <- sub$unsettled_ha / (sub$sustained_ha + sub$endpoint_ha + sub$unsettled_ha)
ramp <- grDevices::colorRampPalette(c(pal[["stable"]], pal[["flicker"]]))(100)
plot(sub$x, sub$y, pch = 15, cex = 0.28, asp = 1, axes = FALSE, ann = FALSE,
     col = ramp[pmax(1, ceiling(frac * 100))])
graphics::title("Whole floodplain, 1 km cells", cex.main = 1.15, font.main = 1)
graphics::rect(win$reach_xmin, win$reach_ymin, win$reach_xmax, win$reach_ymax,
               border = "black", lwd = 1.4)
graphics::legend("bottomleft", title = "share of this cell's\nchange that never settles",
                 legend = c("none", "half", "all"),
                 fill = c(ramp[1], ramp[50], ramp[100]), bty = "n", cex = 0.95,
                 title.cex = 0.9, title.adj = 0)

reach <- terra::unwrap(art$reach)
keys <- c("stable", "break_sustained", "break_endpoint", "flicker", "stable_flicker")
# value-keyed here too: fig_cat codes 0-4 in the order of `keys`
terra::coltab(reach, layer = 1) <- data.frame(value = 0:4, col = unname(pal[keys]))
terra::plot(reach, legend = FALSE, axes = FALSE, mar = NA,
            main = "4 km reach, 10 m cells", cex.main = 1.15, font.main = 1)
# topleft: the reach runs bottom-left to top-right, so this is the empty corner.
# At bottom-left the legend sat on top of the floodplain.
graphics::legend("topleft", legend = unname(lab[keys]), fill = unname(pal[keys]),
                 bty = "n", cex = 0.95)
graphics::par(op)
```

Across the whole Bulkley floodplain, **`r b$pct_sustained`%** of the changed area is a sustained
switch. **`r b$pct_endpoint`%** turns on a single endpoint year, and **`r b$pct_flicker`%** never
settles. Separately, **`r format(round(b$stable_flicker_ha), big.mark = ",")` ha** —
`r b$pct_stable_flicker_of_valid`% of the floodplain — flickers between classes while reading
identically in 2017 and 2023.

## Does Bulkley generalise?

```{r fig-groups, fig.height = 3.4, fig.cap = "Temporal composition of change in the four watershed groups that publish a complete 2017-2023 series, for all change (left) and for tree loss alone (right). Tree loss is every Trees to non-Trees transition except Trees to Clouds, with no patch-size threshold. Four groups is not a sample; the panel shows that the ordering holds and the magnitude moves."}
keys3 <- c("break_sustained", "break_endpoint", "flicker")
all_ch <- data.frame(
  group = rep(grp$group, 3),
  category = rep(keys3, each = nrow(grp)),
  pct = c(grp$pct_sustained, grp$pct_endpoint, grp$pct_flicker),
  panel = "All change")
tl <- tree[tree$class_set == "trees_to_non_trees_excl_clouds", ]
tl <- data.frame(group = tl$group, category = tl$category, pct = tl$pct_of_set,
                 panel = "Tree loss")
# derived, not typed: the prose below quotes both
tree_bulk <- sum(tree$area_ha[tree$group == "bulk" &
                              tree$class_set == "trees_to_non_trees_excl_clouds"])
sus <- function(cs) {
  x <- tree[tree$class_set == cs & tree$category == "break_sustained", ]
  stats::setNames(x$pct_of_set, x$group)
}
shift <- sus("trees_to_non_trees_excl_clouds_water") - sus("trees_to_non_trees_excl_clouds")
dd <- rbind(all_ch, tl)
dd$category <- factor(dd$category, levels = rev(keys3), labels = rev(unname(lab[keys3])))
dd$group <- factor(dd$group, levels = rev(grp$group), labels = rev(unname(group_name[grp$group])))

ggplot2::ggplot(dd, ggplot2::aes(x = pct, y = group, fill = category)) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::facet_wrap(~panel) +
  ggplot2::scale_fill_manual(values = stats::setNames(unname(pal[keys3]), unname(lab[keys3])),
                             breaks = unname(lab[keys3])) +
  ggplot2::scale_x_continuous(expand = c(0, 0), breaks = c(0, 25, 50, 75)) +
  ggplot2::labs(x = "% of changed area", y = NULL, fill = NULL) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "bottom", panel.grid.major.y = ggplot2::element_blank(),
                 panel.spacing.x = grid::unit(1.4, "lines"))
```

The sustained share runs from `r min(grp$pct_sustained)`% to `r max(grp$pct_sustained)`% across
the four groups, and the unsettled share from `r min(grp$pct_flicker)`% to
`r max(grp$pct_flicker)`%. The magnitude moves, and in every group the unsettled share is the
largest of the three. It is not a fixed ordering below that: sustained change is the smallest
category in three groups but not in NECR, where endpoint-only change is slightly smaller.

For tree loss the picture is the same shape but shifted: sustained shares of
`r paste0(round(min(tl$pct[tl$category == "break_sustained"])), "-", round(max(tl$pct[tl$category == "break_sustained"])), "%")`.
Shares are quoted rather than hectares because the two are not equally robust — excluding Trees to
Water alone moves the sustained share by up to `r round(max(shift))` points, which is why the class
set is stated in the caption.

> **Two totals circulate for Bulkley tree loss, and they measure different populations.** The
> published item property `gross_loss_ha` is 1,565.1 ha: patches, after a 1 ha sieve and a
> sub-basin clip. The pixel-level total behind the shares above is
> `r format(round(tree_bulk, 1), big.mark = ",", nsmall = 1)` ha, with no sieve and no clip. The
> ratio between them is the sieve, not a disagreement — it was reconciled separately.

## What this means for a quoted hectare

```{r tbl-groups}
tb <- data.frame(
  Group = unname(group_name[grp$group]),
  `Floodplain (ha)` = round(grp$valid_ha), `Change 2017-2023 (ha)` = round(grp$changed_ha),
  `Sustained (%)` = grp$pct_sustained, `Endpoint-only (%)` = grp$pct_endpoint,
  `Unsettled (%)` = grp$pct_flicker,
  `Stable but flickering (ha)` = round(grp$stable_flicker_ha),
  check.names = FALSE)
knitr::kable(tb, format.args = list(big.mark = ","),
             caption = "Exact values behind this article. Every cell is a column of a CSV shipped with the package and written by a committed script.")
```

A hectare figure from a two-year comparison is not wrong, but it answers a narrower question than
it appears to: *how much area carried a different label in the second year*. On these four
floodplains, between a fifth and a third of that area is change that stayed put for at least two
years on each side. Where the question is durable land-cover change, the sustained share is the
number to reach for — and where it is disturbance extent in a particular year, the endpoint-only
share matters, because a genuine final-year change lives there.

## What this does not support

Four groups is not a sample, and nothing here is a statistical inference; the four are simply the
groups that publish a complete series. Two of them were produced through a different processing
path than the other two and return lower unsettled fractions, so processing and landscape cannot
be separated at this size. And an endpoint-only switch is not evidence of a labelling error: a
real change in the last year of the series produces exactly the same signature, and this method
cannot tell the two apart.

```{r wordcount, include = FALSE}
f <- knitr::current_input()
if (!is.null(f)) {
  x <- readLines(f, warn = FALSE)
  y <- utils::tail(x, -utils::head(which(x == "---"), 2)[2])   # drop the YAML header
  fence <- cumsum(grepl("^```", y)) %% 2 == 1
  y <- y[!fence & !grepl("^```", y)]
  y <- gsub("`r [^`]*`", "", y)          # inline code is not body prose
  y <- y[!grepl("^\\[\\^|^\\s*$", y)]
  n <- sum(lengths(strsplit(trimws(y), "\\s+")))
  message("body prose: ", n, " words")
  if (n > 1000) stop("body prose is ", n, " words; the cap is 1000")
}
```
