---
title: "What a Land-Cover Change Figure Is Made Of"
output: bookdown::html_document2
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  echo = FALSE,
  message = FALSE,
  warning = FALSE,
  fig.width = 7,
  dpi = 150
)
```

```{r load}
library(drift)
d <- function(f) file.path(system.file("extdata", "temporal-composition", package = "drift"), f)

grp <- utils::read.csv(d("summary_groups.csv"), stringsAsFactors = FALSE)
tree <- utils::read.csv(d("summary_treeloss_temporal.csv"), stringsAsFactors = FALSE)
grid <- utils::read.csv(d("bulk_grid_1km.csv"), stringsAsFactors = FALSE)
win <- utils::read.csv(d("bulk_window.csv"), stringsAsFactors = FALSE)
art <- readRDS(d("bulk_window.rds"))
wid <- utils::read.csv(d("summary_patch_widths.csv"), stringsAsFactors = FALSE)
cor <- utils::read.csv(d("summary_corridor.csv"), stringsAsFactors = FALSE)
corc <- utils::read.csv(d("summary_corridor_class.csv"), stringsAsFactors = FALSE)
sv <- utils::read.csv(d("summary_patch_sieve.csv"), stringsAsFactors = FALSE)
ab <- utils::read.csv(d("summary_patch_area_bands.csv"), stringsAsFactors = FALSE)
slv <- utils::read.csv(d("bulk_slivers.csv"), stringsAsFactors = FALSE)
slr <- readRDS(d("bulk_slivers.rds"))

# colours come from the gq registry, so no hex literal appears in this article
reg <- gq::gq_reg_merge(gq::gq_reg_main(),
                        gq::gq_reg_custom(system.file("cartography", "drift_temporal.csv",
                                                      package = "drift")))
tcl <- gq::gq_tmap_classes(reg$layers$temporal_category)
pal <- tcl$values                      # named by class_value, in class order
lab <- stats::setNames(tcl$labels, names(pal))
# a malformed row in the registry CSV adds a phantom layer keyed by its own text
# and gq accepts it without a warning, so assert the class set rather than
# assuming the file parsed as intended
stopifnot(identical(names(pal),
                    c("stable", "break_sustained", "break_endpoint", "unsettled", "stable_flicker")))

b <- grp[grp$group == "bulk", ]
# published watershed-group codes. Only BULK is named in prose, because only
# its full name is established here; inventing the other three would be a fact
# with no source.
group_name <- stats::setNames(toupper(grp$group), grp$group)
```

A published floodplain land-cover change layer is a comparison of two years. For the Bulkley
floodplain (BULK) it says that **`r format(round(b$changed_ha), big.mark = ",")` ha** — about
`r round(b$pct_changed_of_valid)`% of the mapped floodplain — carried a different class in 2023
than in 2017. That number is the input to most things anyone wants to ask.

This article is about what is inside it. It is not about *why* land cover changed: attributing a
patch to a dated fire or cutblock is a separate stage of the same production run, and its results
are written up separately. What follows is the prior question — whether the change a two-year
comparison reports is a change that stayed.

## Looking at every year, not just the ends

The published series carries a classified raster for **every** year from 2017 to 2023, not only
the two endpoints. So each pixel can be read as a seven-year sequence rather than a before-and-
after pair, and sorted by how it behaves:

- **Sustained change** — the class switches once, and holds for at least two years on each side.
- **Endpoint-only change** — the class switches once, but one endpoint year is the only year that
  differs. A real change in the final year of the series looks exactly like this, so it is not the
  same thing as an error.
- **Unsettled** — the class changes more than once and never settles.

A fourth group matters too: pixels that flicker between classes but happen to hold the *same*
class in 2017 and 2023. A two-year comparison cannot see them at all.

```{r fig-patch, fig.height = 4.4, fig.cap = paste0("One ", round(win$area_ha, 2), " ha patch of the Bulkley floodplain mapped as Trees in 2017 and Rangeland in 2023, shown for each year of the series. The outline is the patch; classes are Esri IO LULC. The patch was selected by rule, not by eye: of ", format(win$n_candidates_all, big.mark = ","), " change patches, ", win$n_candidates_transition, " are Trees to Rangeland, ", win$n_candidates_area, " of those are 1 to 4 ha, and ", win$n_candidates_balanced, " of those carry all three temporal categories at 10% or more of their cells. This is the most evenly split of the ", win$n_candidates_balanced, ".")}
lulc <- terra::unwrap(art$lulc)
ct <- dft_class_table("io-lulc")
present <- sort(unique(stats::na.omit(as.vector(terra::values(lulc)))))
ct <- ct[match(present, ct$code), ]
cols <- ct$color

# A colour TABLE, keyed by cell value. `type = "classes"` with a levels/col pair
# maps each layer's own sorted unique values onto the vector positionally, so a
# year that happens to contain no Snow/Ice shifts every class after it -- 2023's
# Rangeland drew in Snow/Ice's blue. Five of the seven years are missing a class,
# so five of seven panels were wrong, and only reading the picture showed it.
tab <- data.frame(value = ct$code, col = ct$color)
for (i in seq_len(terra::nlyr(lulc))) terra::coltab(lulc, layer = i) <- tab

op <- graphics::par(mfrow = c(2, 4), mar = c(0.4, 0.4, 1.6, 0.4), oma = c(0, 0, 0, 0))
for (i in seq_len(terra::nlyr(lulc))) {
  terra::plot(lulc[[i]], legend = FALSE, axes = FALSE, mar = NA, main = names(lulc)[i])
  plot(sf::st_geometry(art$patch), add = TRUE, border = "black", lwd = 1.6)
}
graphics::plot.new()
graphics::legend("center", legend = ct$class_name, fill = cols, bty = "n", cex = 0.95)
graphics::par(op)
```

The patch above is mapped as Trees in 2017 and Rangeland in 2023, so the
published layer counts all `r round(win$area_ha, 2)` ha of it as tree loss. Reading across the
years, only part of it behaves that way.

## The same split across a whole floodplain

```{r fig-map-overview, fig.height = 4.2, fig.cap = "Bulkley floodplain, aggregated to 1 km cells and binned by the share of that cell's changed area that never settles. The box marks the reach shown in the next figure. The floodplain is the drawn area — there is no basemap, so nothing outside the mapped extent is shown."}
sub <- grid[grid$sustained_ha + grid$endpoint_ha + grid$unsettled_ha > 0, ]
frac <- sub$unsettled_ha / (sub$sustained_ha + sub$endpoint_ha + sub$unsettled_ha)
# A raster, not a point cloud: the cells are a regular 1 km grid, so plotting
# them as points meant guessing a `cex` that left gaps at one size and overlap
# at another. And FIVE BINS rather than a continuous ramp -- a 100-step ramp
# between two light registry colours is unreadable at this size, and the
# question a reader has ("is this cell mostly settled or mostly not?") is
# ordinal anyway.
cuts <- c(0, 0.2, 0.4, 0.6, 0.8, 1)
bin_lab <- c("0-20%", "20-40%", "40-60%", "60-80%", "80-100%")
gr <- terra::rast(data.frame(x = sub$x, y = sub$y,
                             v = as.integer(cut(frac, cuts, include.lowest = TRUE))),
                  # bulk_window.csv stores the bare EPSG code, not a CRS string;
                  # terra reads a bare integer as an empty SRS and errors
                  type = "xyz", crs = paste0("EPSG:", win$crs))
bins <- grDevices::colorRampPalette(c(pal[["stable"]], pal[["unsettled"]],
                                      "grey15"))(length(bin_lab) + 1)[seq_along(bin_lab)]
# The registry's temporal colours are all light -- they were chosen to sit under
# black outlines as category fills, not to carry an ordered scale. Ramping
# stable -> unsettled alone spans 29 points of luminance, which at this cell
# size is unreadable; carrying the ramp on into a dark neutral spans 54 and the
# bins separate. `grey15` is a neutral endpoint, not a fifth category colour, so
# no registry value is overridden and no hex literal enters this file.
terra::coltab(gr, layer = 1) <- data.frame(value = seq_along(bin_lab), col = bins)
op <- graphics::par(mar = c(0.4, 0.4, 0.4, 0.4))
terra::plot(gr, legend = FALSE, axes = FALSE, mar = NA)
e <- terra::ext(gr)
graphics::rect(win$reach_xmin, win$reach_ymin, win$reach_xmax, win$reach_ymax,
               border = "black", lwd = 2)
graphics::text(win$reach_xmax, win$reach_ymax, "reach below", pos = 4, cex = 0.85, offset = 0.35)
# EXPLICIT user coordinates, not a keyword position: terra::plot() leaves the
# device in a state where a keyword-placed legend lands outside the visible
# region and silently draws nothing -- the first version of this figure
# published with no legend at all.
# Position measured, not guessed: binning the occupied 1 km cells onto a 10x10
# grid of the extent, the blocks x 0-0.6 by y 0-0.2 and x 0.6-1 by y 0.6-1 are
# the two that are entirely empty. The top-right one is 0.4 x 0.4 and the
# bottom-left 0.6 x 0.2; a five-entry legend with a two-line title needs about
# 0.32 of the height, so only the top-right one fits it. Three earlier
# placements -- the "bottomleft" keyword, the top-left corner, and bottom-left
# by coordinate -- landed on floodplain, on floodplain, and clipped off the
# bottom of the device respectively.
graphics::legend(x = e[1] + 0.62 * (e[2] - e[1]), y = e[4],
                 title = "share of this cell's change\nthat never settles",
                 legend = bin_lab, fill = bins, bty = "n", cex = 0.85,
                 title.cex = 0.85, title.adj = 0, xpd = NA)
graphics::par(op)
```

```{r fig-map-reach, fig.height = 3.8, fig.cap = "The same floodplain at the mapped 10 m resolution, over a 4 km reach — the box in the previous figure. Each cell carries its own temporal category rather than a share."}
reach <- terra::unwrap(art$reach)
keys <- c("stable", "break_sustained", "break_endpoint", "unsettled", "stable_flicker")
# value-keyed here too: fig_cat codes 0-4 in the order of `keys`
terra::coltab(reach, layer = 1) <- data.frame(value = 0:4, col = unname(pal[keys]))
op <- graphics::par(mar = c(0.4, 0.4, 0.4, 0.4))
terra::plot(reach, legend = FALSE, axes = FALSE, mar = NA)
# topleft: the reach runs bottom-left to top-right, so this is the empty corner.
# At bottom-left the legend sat on top of the floodplain.
graphics::legend("topleft", legend = unname(lab[keys]), fill = unname(pal[keys]),
                 bty = "n", cex = 0.95)
graphics::par(op)
```

Across the whole Bulkley floodplain, **`r b$pct_sustained`%** of the changed area is a sustained
switch. **`r b$pct_endpoint`%** turns on a single endpoint year, and **`r b$pct_unsettled`%** never
settles. Separately, **`r format(round(b$stable_flicker_ha), big.mark = ",")` ha** —
`r b$pct_stable_flicker_of_valid`% of the floodplain — flickers between classes while reading
identically in 2017 and 2023.

## Does Bulkley generalise?

```{r fig-groups, fig.height = 3.4, fig.cap = "Temporal composition of change in the four watershed groups that publish a complete 2017-2023 series, for all change (left) and for tree loss alone (right). Tree loss is every Trees to non-Trees transition except Trees to Clouds, with no patch-size threshold. Four groups is not a sample; the panel shows that the ordering holds and the magnitude moves."}
keys3 <- c("break_sustained", "break_endpoint", "unsettled")
all_ch <- data.frame(
  group = rep(grp$group, 3),
  category = rep(keys3, each = nrow(grp)),
  pct = c(grp$pct_sustained, grp$pct_endpoint, grp$pct_unsettled),
  panel = "All change")
tl <- tree[tree$class_set == "trees_to_non_trees_excl_clouds", ]
tl <- data.frame(group = tl$group, category = tl$category, pct = tl$pct_of_set,
                 panel = "Tree loss")
# derived, not typed: the prose below quotes both
tree_bulk <- sum(tree$area_ha[tree$group == "bulk" &
                              tree$class_set == "trees_to_non_trees_excl_clouds"])
sus <- function(cs) {
  x <- tree[tree$class_set == cs & tree$category == "break_sustained", ]
  stats::setNames(x$pct_of_set, x$group)
}
shift <- sus("trees_to_non_trees_excl_clouds_water") - sus("trees_to_non_trees_excl_clouds")
dd <- rbind(all_ch, tl)
dd$category <- factor(dd$category, levels = rev(keys3), labels = rev(unname(lab[keys3])))
dd$group <- factor(dd$group, levels = rev(grp$group), labels = rev(unname(group_name[grp$group])))

ggplot2::ggplot(dd, ggplot2::aes(x = pct, y = group, fill = category)) +
  ggplot2::geom_col(width = 0.7) +
  ggplot2::facet_wrap(~panel) +
  ggplot2::scale_fill_manual(values = stats::setNames(unname(pal[keys3]), unname(lab[keys3])),
                             breaks = unname(lab[keys3])) +
  ggplot2::scale_x_continuous(expand = c(0, 0), breaks = c(0, 25, 50, 75)) +
  ggplot2::labs(x = "% of changed area", y = NULL, fill = NULL) +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "bottom", panel.grid.major.y = ggplot2::element_blank(),
                 panel.spacing.x = grid::unit(1.4, "lines"))
```

The sustained share runs from `r min(grp$pct_sustained)`% to `r max(grp$pct_sustained)`% across
the four groups, and the unsettled share from `r min(grp$pct_unsettled)`% to
`r max(grp$pct_unsettled)`%. The magnitude moves, and in every group the unsettled share is the
largest of the three. It is not a fixed ordering below that: sustained change is the smallest
category in three groups but not in NECR, where endpoint-only change is slightly smaller.

For tree loss the picture is the same shape but shifted: sustained shares of
`r paste0(round(min(tl$pct[tl$category == "break_sustained"])), "-", round(max(tl$pct[tl$category == "break_sustained"])), "%")`.
Shares are quoted rather than hectares because the two are not equally robust — excluding Trees to
Water alone moves the sustained share by up to `r round(max(shift))` points, which is why the class
set is stated in the caption.

> **Two totals circulate for Bulkley tree loss, and they measure different populations.** The
> published item property `gross_loss_ha` is 1,565.1 ha: patches, after a 1 ha sieve and a
> sub-basin clip. The pixel-level total behind the shares above is
> `r format(round(tree_bulk, 1), big.mark = ",", nsmall = 1)` ha, with no sieve and no clip. The
> ratio between them is the sieve, not a disagreement — it was reconciled separately.

## The shape of a change patch

```{r shape-prep, include = FALSE}
wpick <- function(g, pg, col) wid[wid$group == g & wid$patch_group == pg, col]
sl_n <- sapply(grp$group, function(g) 100 * wpick(g, "sliver", "n_patches") / wpick(g, "all", "n_patches"))
sl_a <- sapply(grp$group, function(g) 100 * wpick(g, "sliver", "area_ha") / wpick(g, "all", "area_ha"))
nf_s <- sapply(grp$group, wpick, pg = "sliver", col = "n_flips_area_wtd")
nf_w <- sapply(grp$group, wpick, pg = "wider", col = "n_flips_area_wtd")
bf_s <- sapply(grp$group, wpick, pg = "sliver", col = "break_frac_area_wtd")
bf_w <- sapply(grp$group, wpick, pg = "wider", col = "break_frac_area_wtd")
# collapse a degenerate range: with four groups the endpoints often round to the
# same value, and "2-2 cells" reads as a typo rather than as agreement
rng <- function(x, dp = 0) {
  v <- formatC(range(x), format = "f", digits = dp)
  if (v[1] == v[2]) v[1] else paste(v, collapse = "-")
}
stopifnot(all(nf_s > nf_w), all(bf_s < bf_w))          # the unsieved claim, asserted

s0 <- sv[sv$sieve_ha == 0, ]; s1 <- sv[sv$sieve_ha == 1, ]
med_cells <- s0$sliver_area_ha_median * 100          # 0.01 ha per cell at 10 m
# the area-controlled comparison, in the bands where width is actually variable
abv <- ab[ab$area_band %in% c("0.2-0.5", "0.5-1") & ab$n_sliver >= 30 & ab$n_wider >= 30, ]
n_rev <- sum(abv$break_frac_sliver > abv$break_frac_wider)
sm <- ab[ab$area_band == "0.02-0.1", ]
n_wider_small <- sum(sm$n_wider); n_small <- sum(sm$n_patches)
stopifnot(nrow(abv) >= 6, n_rev > nrow(abv) / 2,      # it reverses in most of them
          all(sm$pct_sliver >= 99.9))                 # not 100: necr has three
```

Change patches are overwhelmingly narrow: between `r rng(sl_n)`% of them have an effective width
below one and a half pixels — the measure is `2 x area / perimeter`, so small compact blobs count
as well as one-cell strips — and they carry only `r rng(sl_a)`% of the changed area. Unsieved, they
also flicker more than wider patches (`r rng(nf_s, 2)` switches against `r rng(nf_w, 2)`) and
settle less often (`r rng(bf_s, 3)` of their area is a clean break, against `r rng(bf_w, 3)`).

They are also very small. The median sliver is `r rng(med_cells, 0)` cells. That matters, because
the usual conservative move is not to filter on shape at all but to drop small patches by area —
the published change layer already discards everything under a hectare. **That sieve keeps
`r rng(s1$pct_patches_kept, 1)`% of the patches and `r rng(s1$pct_area_kept, 1)`% of the changed
area, and only `r rng(s1$pct_kept_sliver, 1)`% of what survives is a sliver.** So for anyone
working from the sieved layer, this population is already gone; and the sieve is discarding a third
to a half of the changed area to do it.

Holding area fixed instead of sieving is what separates the two ideas, and it undoes the result.
Below a tenth of a hectare all but `r n_wider_small` of `r format(n_small, big.mark = ",")` patches
are slivers, so width distinguishes nothing there. Above
a fifth of a hectare, where narrow and compact patches of the same size both exist in numbers, the
gap does not merely shrink — it **reverses** in `r n_rev` of the `r nrow(abv)` group-and-size cells
where both populations are large enough to compare: narrow patches settle *more* often than compact
ones of the same area. Width was standing in for size.

A second geometric signature asks whether a patch traces a pre-existing boundary between the two
classes rather than cutting across one. It does not generalise either: tracing patches settle less
often than the rest in three groups and slightly more often in the fourth.

```{r fig-slivers, fig.height = 5.6, fig.cap = "Two change patches from the Bulkley floodplain, each outlined in black, with 250 m of context. Top: a sliver at the margin of permanent water. Bottom: a sliver on the edge of a clearing, 13 km from any permanent water. Left and centre are the mapped land cover in the first and last year; right is the temporal category of every cell in the window. Both were chosen by a stated rule from the 0.2-0.6 ha band and are recorded, with that rule, in the figure data."}
ct <- dft_class_table("io-lulc")
ctab <- data.frame(value = ct$code, col = ct$color)
tkeys <- c("stable", "break_sustained", "break_endpoint", "unsettled", "stable_flicker")
tcab <- data.frame(value = 0:4, col = unname(pal[tkeys]))
setting_lab <- c(water = "beside permanent water", boundary = "at a dry class boundary")
# a legend strip across the bottom, not a fourth column: the panels are square
# and shrinking them to make room would cost the cell detail the figure exists
# to show
graphics::layout(matrix(c(1, 2, 3, 4, 5, 6, 7, 7, 7), nrow = 3, byrow = TRUE),
                 heights = c(1, 1, 0.42))
op <- graphics::par(mar = c(0.3, 0.3, 1.7, 0.3))
present <- integer()
for (nm in names(setting_lab)) {
  z <- slr[[nm]]
  for (key in c("from", "to", "cat")) {
    r <- terra::unwrap(z[[key]])
    if (key != "cat") present <- union(present, stats::na.omit(unique(terra::values(r))))
    # value-keyed colour table, never a positional levels/col pair
    terra::coltab(r, layer = 1) <- if (key == "cat") tcab else ctab
    ttl <- paste0(setting_lab[[nm]], ": ",
                  c(from = "2017", to = "2023", cat = "temporal category")[[key]])
    terra::plot(r, legend = FALSE, axes = FALSE, mar = NA,
                main = ttl, cex.main = 1.05, font.main = 1)
    plot(sf::st_geometry(z$geom), add = TRUE, border = "black", lwd = 1.8, col = NA)
  }
}
# only the classes these two windows actually contain -- a legend built from the
# whole class table would list seven classes for a figure that shows four
present <- sort(present)
kp <- ct[match(present, ct$code), ]
stopifnot(!anyNA(kp$class_name))
graphics::par(mar = c(0.2, 0.2, 0.2, 0.2))
graphics::plot.new()
graphics::legend("left", horiz = FALSE, ncol = 2, title = "land cover (left, centre)",
                 legend = kp$class_name, fill = kp$color, bty = "n", cex = 1.0,
                 title.cex = 0.95, title.adj = 0)
graphics::legend("right", horiz = FALSE, ncol = 2, title = "temporal category (right)",
                 legend = unname(lab[tkeys]), fill = unname(pal[tkeys]), bty = "n",
                 cex = 1.0, title.cex = 0.95, title.adj = 0)
graphics::par(op)
graphics::layout(1)
```

One is a ribbon along a channel margin and the other an arc on the edge of a clearing thirteen
kilometres from any water, and they are the same object to the width test: one to two cells across,
tracing an interface that was already there. That is the point. Geometry alone does not say whether
a thin patch is a mixed pixel at a water margin, a registration artifact, or a real narrow change —
and the two panels on the right show the settings differ in *when* they move, which is the axis the
shape tests cannot see.

## Where the instability sits

```{r fig-corridor, fig.height = 3.6, fig.cap = "Flicker as a function of distance, in each of the four groups. The purple line measures distance from permanent water — pixels classed Water in all seven years. The blue line is the null: distance from a from-epoch class boundary with no water on either side. Both use the same bands and the same denominator, the scanned cells in that band. The reference band itself is omitted from both, being a definition rather than a measurement."}
lv <- c("0-10", "10-30", "30-50", "50-100", "100-200", "200-500", ">500")
fl <- cor[cor$category_label %in% c("unsettled", "stable_flicker") &
            cor$reference %in% c("water_core", "edge_nonwater") & cor$band_label %in% lv, ]
fl <- stats::aggregate(pct_of_scanned_in_band ~ group + reference + band_label, fl, sum)
fl$band_label <- factor(fl$band_label, levels = lv)
fl$panel <- factor(unname(group_name[fl$group]), levels = unname(group_name[grp$group]))
fl$ref <- factor(ifelse(fl$reference == "water_core", "from permanent water",
                        "from a non-water class boundary"),
                 levels = c("from permanent water", "from a non-water class boundary"))
# two registry colours used as LINE identities; no hex literal enters this file
ln <- stats::setNames(unname(pal[c("unsettled", "break_sustained")]), levels(fl$ref))
ggplot2::ggplot(fl, ggplot2::aes(band_label, pct_of_scanned_in_band, colour = ref, group = ref)) +
  ggplot2::geom_line(linewidth = 0.7) + ggplot2::geom_point(size = 1.1) +
  ggplot2::facet_wrap(~panel, nrow = 1) +
  ggplot2::scale_colour_manual(values = ln, name = NULL) +
  ggplot2::labs(x = "distance (m)", y = "flicker (% of scanned cells in band)") +
  ggplot2::theme_minimal(base_size = 11) +
  ggplot2::theme(legend.position = "bottom",
                 axis.text.x = ggplot2::element_text(angle = 45, hjust = 1, size = 8),
                 panel.spacing = grid::unit(0.9, "lines"))
```

```{r corridor-prep, include = FALSE}
tr <- corc[corc$from_class == "Trees" & corc$category_label %in% c("unsettled", "stable_flicker"), ]
tr <- stats::aggregate(pct_of_band_class ~ group + band_label, tr, sum)
tnear <- tr$pct_of_band_class[tr$band_label == "0-10"]
tfar <- tr$pct_of_band_class[tr$band_label == ">500"]
enear <- sapply(grp$group, function(g) sum(fl$pct_of_scanned_in_band[fl$group == g & fl$band_label == "0-10" & fl$reference == "edge_nonwater"]))
efar <- sapply(grp$group, function(g) sum(fl$pct_of_scanned_in_band[fl$group == g & fl$band_label == ">500" & fl$reference == "edge_nonwater"]))
wfar <- sapply(grp$group, function(g) sum(fl$pct_of_scanned_in_band[fl$group == g & fl$band_label == ">500" & fl$reference == "water_core"]))
# the claims in the prose below, asserted rather than read off a figure
stopifnot(length(tnear) == nrow(grp), all(tnear > tfar), all(enear > efar),
          all(wfar > efar),                       # water sits above the null throughout
          all((enear / efar) > (rng_w <- sapply(grp$group, function(g)
            sum(fl$pct_of_scanned_in_band[fl$group == g & fl$band_label == "0-10" & fl$reference == "water_core"]))
            / wfar)))                              # and the null falls further
```

The reach map above hints that the unstable cells trace the channel, and they do. Within the Trees
class alone — a Trees pixel can never be part of the water reference, so the comparison is not
measuring its own definition — flicker runs `r rng(tnear, 1)`% in the first ten metres beside
permanent water against `r rng(tfar, 1)`% more than five hundred metres away, falling in every band.

The null settles how to read that. Against a from-epoch class boundary with **no water on either
side** the same gradient appears, and falls further: `r rng(enear, 1)`% in the first ten metres to
`r rng(efar, 1)`% beyond five hundred. The water profile sits above it throughout and levels off near
`r rng(wfar, 1)`% — what a cell far from the river but near some other boundary looks like. So a
water margin is the most unstable edge, not a different kind of thing. Flicker
concentrates at class boundaries, and the channel is simply the longest, most sinuous,
highest-contrast boundary a floodplain has — the corridor is an edge effect in a river's shape.

Nor does the near-channel change look like a channel that moved. A migrating bank would date its
switches progressively outward; instead the mean year of a sustained break beside permanent water
is *later* than one far away, in all four groups — the opposite of a walk.

## What this means for a quoted hectare

```{r tbl-groups}
tb <- data.frame(
  Group = unname(group_name[grp$group]),
  `Floodplain (ha)` = round(grp$valid_ha), `Change 2017-2023 (ha)` = round(grp$changed_ha),
  `Sustained (%)` = grp$pct_sustained, `Endpoint-only (%)` = grp$pct_endpoint,
  `Unsettled (%)` = grp$pct_unsettled,
  `Stable but flickering (ha)` = round(grp$stable_flicker_ha),
  check.names = FALSE)
knitr::kable(tb, format.args = list(big.mark = ","),
             caption = "Exact values behind this article. Every cell is a column of a CSV shipped with the package and written by a committed script.")
```

A hectare figure from a two-year comparison is not wrong, but it answers a narrower question than
it appears to: *how much area carried a different label in the second year*. On these four
floodplains, between a fifth and a third of that area is change that stayed put for at least two
years on each side. Where the question is durable land-cover change, the sustained share is the
number to reach for — and where it is disturbance extent in a particular year, the endpoint-only
share matters, because a genuine final-year change lives there.

## What this does not support

Four groups is not a sample, and nothing here is a statistical inference; the four are simply the
groups that publish a complete series. Two of them were produced through a different processing
path than the other two and return lower unsettled fractions, so processing and landscape cannot
be separated at this size. And an endpoint-only switch is not evidence of a labelling error: a
real change in the last year of the series produces exactly the same signature, and this method
cannot tell the two apart.

The spatial results carry their own limits. Distance is measured from permanent water rather than
a mapped stream network, and in one group two thirds of the floodplain is a regulated lake. The stable share of a near band is biased downward by construction, because every
permanent-water pixel sits in the reference rather than in the bands — which is why the number
above is taken within Trees. And an edge effect is not a cause: nothing separates a boundary a
classifier renders inconsistently from one that genuinely moved, and the break dates rule out only
the steady outward walk.

```{r wordcount, include = FALSE}
f <- knitr::current_input()
if (!is.null(f)) {
  x <- readLines(f, warn = FALSE)
  y <- utils::tail(x, -utils::head(which(x == "---"), 2)[2])   # drop the YAML header
  fence <- cumsum(grepl("^```", y)) %% 2 == 1
  y <- y[!fence & !grepl("^```", y)]
  y <- gsub("`r [^`]*`", "", y)          # inline code is not body prose
  y <- y[!grepl("^\\[\\^|^\\s*$", y)]
  n <- sum(lengths(strsplit(trimws(y), "\\s+")))
  message("body prose: ", n, " words")
  # 1000 at #66; 1200 at #73 for the shape and location sections; 1500 once the
  # patch-size sieve and the two worked sliver examples were added, because the
  # sieve result is what a reader of the published (already sieved) layer needs
  # and it cannot be stated in a clause. A variable, not two literals -- the
  # number used to appear twice on one line and changing one left the message
  # lying about the other.
  cap <- 1500
  if (n > cap) stop("body prose is ", n, " words; the cap is ", cap)
}
```
